---
title: "Personal Activity Monitors"
author: "Cary Correia"
date: "December 16, 2014"
output: html_document
---

###Executive Summary

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data 
about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – 
a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in 
their behavior, or because they are tech geeks. 

The goal of this project is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants and 
to build a model that can predict the manner in which they performed barbell lifts correctly (A-E).
More information is available from the website here: http://groupware.les.inf.puc-rio.br/har.
 
The analytical flow was as follows:  1) download datasets; Training and Validation, (2) Clean and prep both datasets,
(3) Figure out what data is not important with a screen model (4) Create a tuned model with a reduced variable set (5) Test the new model on the 'testing' dataset and (6) Use the final model to predict the values in the sample test data (20 rows)

Final results showed a final model accuracy of 99.04% with an out of sample error equal to 0.96%.  These results were verified with cross-validation via a training (60% of sample) and testing (40%) set of data.  Final prediction on our sample of 20 test scenarios yielded 100% (see report)

####Section 1) download datasets; Training and Testing

```{r}
#Prep....setup libraries and set.seed
library(caret);library(randomForest); set.seed(12142014)    

#Set the train and test url's
data.url<-'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv'        
test.url<-'http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv'

#Get and save the data
data       <-read.csv(url(data.url), header=TRUE, sep=",", fill=TRUE, na.strings=c("NA", "", " "))               
validation <- read.csv(url(test.url), header=TRUE, sep=",", na.strings="NA")
```

####Section 2) Clean and prep both datasets
```{r}
# Create a function to detect which column has lots of NA's - count #NA's / totalRows
na.test <- function(x,y) {
    w<-sapply(x, function(x)length(which(!is.na(x)))/length(x)*100)          # calculate %-age of rows that are NA in each col
    z<<-as.data.frame(sapply(w, function(x)if(x>=y) keep<-x else keep<-NA))  # creates a vector keep that has columns we will keep
    names(z)<<-c("value")                                                    # fixes the header of the z table
}
na.test(data, 99)                                                       # input-> dataset + threshold for non-NA's
keep<-as.data.frame(row.names(na.omit(z))); names(keep)<-'label'
keep<-as.character(keep$label[8:60])                                    # output-> vector of all rows to keep

data<-data[,names(data) %in% keep]                                      # downselect the variables we want 
nzero<-nearZeroVar(data, saveMetrics=TRUE)                              # check all remaining vars to see if they are near zero
                                                                        # no non zero's were found- so nothing else to remove
inTrain    <-createDataPartition(data$classe, p=.20, list = FALSE)      # create inTrain string for data partitions   
training   <-data[inTrain,]                                             # this sample for 1st round eval of variables is low=3.9K
```

####Section 3) Figure out what data is not important
```{r}
library(doMC)                                                                   # this code enables parllel processing
registerDoMC(cores = 4)                                                         # All subsequent models are then run in parallel

screen<-randomForest(classe ~., data=training, importance=TRUE, proximity=TRUE) # 1st screen--> 3.9K obs with 53 variables....
vi<-importance(screen)                                                          # Create varImportance matrix
vi_vals<-as.data.frame(vi[,1:5])                                                # Save raw scores-> Higher is better
top_vi<-as.data.frame(apply(vi_vals, 2, rank))                                  # Convert scores to rank
top_vi$total<-rowSums(top_vi); top_sort<-top_vi[order(-top_vi$total),]          # Sum ranks-> Hi totals = more importance
```

The following chart illustrates the variables from most important to least.  We will now introduce a cutoff to shrink the number of variables in the next model. Note: the variable importance chart values were replaced with force rank scores.  By adding across the scores it was easy to assess which variables were the most valuable in our model:
```{r, echo=FALSE}
top_sort
```

####Section (4) Create a tuned model with a reduced variable set-> but only use the variables that pass section (3)
```{r}
# cut off the bottom of the variables....include these for the actual full model run
keep2<-as.character(row.names(top_sort[top_sort$total>=88,]))                   # Take only variables >=88
keep2[37]<-"classe"                                                             # Add back in the "classe" output

data2<-data[,names(data) %in% keep2]                                            # Build the dataset with just the top values
inTrain2    <-createDataPartition(data2$classe, p=.60, list = FALSE)            # create inTrain string for data partitions   (60:40 split)
training2   <-data2[inTrain2,]                                                  # split data into training and testing 
testing2    <-data2[-inTrain2,]

model<-train(classe ~., data=training2, method='rf', importance=TRUE)           # final model: random forest--> 12K obs with 36 variables
vi2<-varImp(model)
```

####Section 5) Test the new model on the 'testing' dataset
```{r}
accuracy<-confusionMatrix(testing2$classe, predict(model, testing2))            # apply model to testing data
accuracy                                                                        # print out accuracy (99.06%)
```
The accuracy of the model is `r acc<-round(accuracy$overall[1]*100, 2);acc` and the out of sample error is `r 100-acc`

#####Section 6) Use the finalModel to predict the values in the sample test data (20 rows)
```{r}
# prep validation data
keep2[37]<- "problem_id"                                                        # fix keep2 vector and replace classe with prob ID
test_final<-validation[,names(validation) %in% keep2]                           # prep the test data to downselect the model variables
f.predict<-as.character(predict(model, test_final))                             # create a final prediction vector w. 20 answers
```
The final predicted string for the 20 test samples was `r f.predict`.  
This list was submitted and all 20 were correct.


####Appendix

Full Variable Importance Matrix for all outcomes 1-5:
```{r}
plot(vi2, asp=4)
```

